---
title: "PPA Midterm"
author: "RVM"
date: "2024-03-20"
output: html_document:
  keep_md: yes
  toc: yes
  theme: flatly
  toc_float: yes
  code_folding: hide
  number_sections: no
  fontsize: 12pt
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE)
```

Property Value Prediction Model 
have variables, may wanna gather more for the model
predictive model of price incorporating structural and locational characteristics 

- summary statistics w variable descriptions
- correlation matrix
- 4 home price correlation scatterplots
- 1 map of dep variable (sale price)
- 3 maps of 3 most interesting independent variables (factors we choose)
- testing and training dataset???
- discussion of performance
- plot of predicted prices as function of observed prices
- map of residuals 

## Introduction 
Clearly states the goal of the analysis	
Provides a brief overview of the methods for data collection

## Data Description
Presents a comprehensive table of summary statistics	
Variable descriptions are included and sorted by category

## Exploratory Data Analysis 
Correlation matrix is provided	
Four home price correlation scatterplots are included	
Map of the dependent variable (sale price) is presented	
Three maps of independent variables are included

## Modeling 
5 points to hit here

## Conclusion
Summarizes findings and discusses limitations	
Identifies potential areas for further model improvement


```{r setting up the packages}
if (!require(pacman)){install.packages("pacman"); library(pacman)}
p_load (sf, tidyverse, knitr, kableExtra, rmarkdown, tidycensus, dplyr, scales)

```

```{r}
census_api_key("b3eda1fa84dde3c5ad443fd407d48f2584ab2726", overwrite = TRUE)

kingsCounty <- read.csv("kc_house_data.csv")
st_crs(kingsCounty)

seattleSchools <- read.csv("Seattle_Public_Schools_Sites_2023-2024.csv")
seattleTreeCanopy <- read.csv("Seattle_Tree_Canopy_2016_2021_RSE_Census_Tracts.csv")
seattleSidewalks <- read.csv("Sidewalks-API_20240326.csv")

```


```{r}
#filtering out data that's not seattle

seattleData <- kingsCounty %>%
  filter(zipcode %in% c("98101","98102","98103","98104","98105", "98106","98107", "98108","98109","98111","98112","98113",'98114","98115"."98116","98117","98118","98119","98120","98121","98122","98123","98124","98125","98126","98131","98132',"98133","98134","98135","98136","98138","98144","98145","98146","98148","98154","98155","98158","98160","98161","98164","98166","98168","98171","98174","98177","98178","98188","98198","98199"))


seattleData <- seattleData %>%
  select(price, condition, sqft_lot, lat, long, yr_built, yr_renovated, zipcode)
seattleDisability <- seattleTreeCanopy %>%
  select(PCT_ADULT_WITH_DISABILITIES)

#summary statistics should ideally combine the different datasets 
#want the summary stats to cover internal (condition, lot size, year built/year renovated) and external (disabilities and sidewalk conditions)
summary(seattleData)


```


