---
title: "PPA Midterm"
author: "RVM"
date: "2024-03-20"
output: html_document:
  keep_md: yes
  toc: yes
  theme: flatly
  toc_float: yes
  code_folding: hide
  number_sections: no
  fontsize: 12pt
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE)
```

Property Value Prediction Model 
have variables, may wanna gather more for the model
predictive model of price incorporating structural and locational characteristics 

- summary statistics w variable descriptions
- correlation matrix
- 4 home price correlation scatterplots
- 1 map of dep variable (sale price)
- 3 maps of 3 most interesting independent variables (factors we choose)
- testing and training dataset???
- discussion of performance
- plot of predicted prices as function of observed prices
- map of residuals 

## Introduction 
Clearly states the goal of the analysis	
Provides a brief overview of the methods for data collection

## Data Description
Presents a comprehensive table of summary statistics	
Variable descriptions are included and sorted by category

## Exploratory Data Analysis 
Correlation matrix is provided	
Four home price correlation scatterplots are included	
Map of the dependent variable (sale price) is presented	
Three maps of independent variables are included

## Modeling 
5 points to hit here

## Conclusion
Summarizes findings and discusses limitations	
Identifies potential areas for further model improvement


```{r setting up the packages}
if (!require(pacman)){install.packages("pacman"); library(pacman)}
p_load (sf, tidyverse, knitr, kableExtra, rmarkdown, tidycensus, dplyr, scales)

```

```{r}
census_api_key("b3eda1fa84dde3c5ad443fd407d48f2584ab2726", overwrite = TRUE)

kings_county <- read_csv("kc_house_data.csv")
st_crs(kings_county)
st_transform(kings_county, crs = 4326)

```

