---
title: "PPA Midterm"
author: "Kamya Khandelwal, Revathi Machan"
date: "2024-03-20"
output: html_document:
  keep_md: yes
  toc: yes
  theme: flatly
  toc_float: yes
  code_folding: hide
  number_sections: no
  fontsize: 12pt
editor_options:
  markdown:
    wrap: 72
---
#making changes and trying new things 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE)
```

Property Value Prediction Model 
have variables, may wanna gather more for the model
predictive model of price incorporating structural and locational characteristics 

- summary statistics w variable descriptions
- correlation matrix
- 4 home price correlation scatterplots
- 1 map of dep variable (sale price)
- 3 maps of 3 most interesting independent variables (factors we choose)
- testing and training dataset???
- discussion of performance
- plot of predicted prices as function of observed prices
- map of residuals 

## Introduction 
Clearly states the goal of the analysis	
Provides a brief overview of the methods for data collection

## Data Description
Presents a comprehensive table of summary statistics	
Variable descriptions are included and sorted by category

## Exploratory Data Analysis 
Correlation matrix is provided	
Four home price correlation scatterplots are included	
Map of the dependent variable (sale price) is presented	
Three maps of independent variables are included

## Modeling 
5 points to hit here

## Conclusion
Summarizes findings and discusses limitations	
Identifies potential areas for further model improvement


```{r setting up the packages}
if (!require(pacman)){install.packages("pacman"); library(pacman)}
p_load (sf, tidyverse, knitr, kableExtra, rmarkdown, tidycensus, dplyr, scales, stringr, ggcorrplot)

```

```{r}
census_api_key("b3eda1fa84dde3c5ad443fd407d48f2584ab2726", overwrite = TRUE)
```

# Data Wrangling

Census, OpenData Seattle and provided Dataset 

## Provided Dataset and Internal Variables

The code below is used to import the foundational dataset for our model. This dataset contains information on home sales prices and property characteristics in Seattle, Kings County for the years 2014 and 2015. The model we are developing is designed to make predictions about home prices and utilizes specific property attributes from this dataset to enhance the accuracy of those predictions.

```{r Reading Data, results= 'hide'}

kingsCounty <- read.csv("kc_house_data.csv")
st_crs(kingsCounty)

seattleSchools <- read.csv("Seattle_Public_Schools_Sites_2023-2024.csv")
seattleTreeCanopy <- read.csv("Seattle_Tree_Canopy_2016_2021_RSE_Census_Tracts.csv")
seattleSidewalks <- read.csv("Sidewalks-API_20240326.csv")

```


```{r}
#filtering out data that's not Seattle

seattleData <- kingsCounty %>%
  filter(zipcode %in% c("98101","98102","98103","98104","98105", "98106","98107", "98108","98109","98111","98112","98113",'98114","98115"."98116","98117","98118","98119","98120","98121","98122","98123","98124","98125","98126","98131","98132',"98133","98134","98135","98136","98138","98144","98145","98146","98148","98154","98155","98158","98160","98161","98164","98166","98168","98171","98174","98177","98178","98188","98198","98199"))


seattleData <- seattleData %>%
  select(price, condition, sqft_living, lat, long, yr_built, bedrooms, zipcode, sqft_basement)

seattleData <- seattleData %>%
  filter(yr_built > 0) 

#categorizing based on basement presence
seattleData <- seattleData %>%
  mutate(basementPresent = case_when(
    sqft_basement == 0 ~ 'No Basement',
    sqft_basement >= 1 ~ 'Basement'))

#categorizing based on condition
seattleData <- seattleData %>%
  mutate(conditionType  = case_when(
    condition == '1' ~ 'Very Poor',
    condition == '2' ~ 'Poor',
    condition == '3' ~ 'Average',
    condition == '4' ~ 'Good',
    condition == '5' ~ 'Very Good'))

seattleData <- seattleData %>%
  mutate(houseSize = case_when(
    sqft_living >= 2000 ~ 'Big',
    sqft_living <= 1999 ~ 'Small'))

seattleDisability <- seattleTreeCanopy %>%
  select(PCT_ADULT_WITH_DISABILITIES)


#summary statistics should ideally combine the different datasets 
#want the summary stats to cover internal (condition, lot size, year built/year renovated) and external (disabilities and sidewalk conditions)
summary(seattleData)

```


```{r corr int}
seattleDatacorr <- seattleData %>%
  select(price, condition, sqft_living, yr_built, bedrooms, sqft_basement)

cor(seattleDatacorr)
corr <- round(cor(seattleDatacorr), 1)
ggcorrplot(corr,
           type = "upper",
           lab = TRUE, 
           lab_size = 5,  
           colors = c('#d7191c','#ffffbf','#2c7bb6'),
           title="Correlation Matrix of Housing Dataset", 
           ggtheme=theme_bw)
```

```{r impute missing data}

seattleData <- seattleData %>%
  mutate(bedroomIndex = case_when(
    bedrooms >= 1 ~ 1, 
    bedrooms < 1 ~ 0))

seattleData$bedroomIndex[is.na(seattleData$bedroomIndex)] <- 0

lm(bedrooms ~ sqft_living, data = seattleData)
for(i in 1:nrow(seattleData))
{
  if(seattleData$bedroomIndex[i] == 0)
  {seattleData$bedrooms[i] = 1.9054627 +  0.0007142*seattleData$sqft_living}
}

seattleData$bedrooms <- round(seattleData$bedrooms, digits = 0)


```

```{r creating new variables}

seattleData <- seattleData %>%
  mutate(pricePerFt2 = (price/sqft_living)) #might need to do the linear regression bathroom and rooms for this stuff

seattleData <- seattleData %>%
  mutate(buildingAge = (2015 -(yr_built)))
```


###Census Data Collection

```{r}
acsVariables2015 <- load_variables(2015, "acs5", cache = TRUE)

# 2021, A

# Retrieve ACS data for Philadelphia tracts in 2020
kcTracts <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",   # Total Population
    "B19013_001",   # Median Household Income
    "B25058_001",   # Median Rent
    "B25008_002",   # Owner-Occupied Units
    "B25008_003",   # Renter-Occupied Units
    "B07001_032",   # Same House 75 Years Ago
    "B07001_017",   # Same House 1 Year Ago
    "B25088_003",   # Median Selected Monthly Owner Costs (homes without a mortgage)
    "B25088_002",   # Median Selected Monthly Owner Costs (homes with a mortgage)
    "B15003_022",   # Educational Attainment: Bachelor's Degree
    "B17001_002",   # Percentage of Population Below the Poverty Level
    "B28002_004",   # Percentage of Housing Units with High-Speed Internet
    "B25044_003",   # Percentage of Housing Units with No Vehicle Available
    "B02001_002",   # Race and Ethnicity: White Alone
    "B02001_003",   # Race and Ethnicity: Black or African American Alone
    "B03001_003"    # Hispanic or Latino Origin of Population
  ),
  year = 2021,
  state = "WA",
  county = "King",
  geometry = TRUE,
  output = "wide"
)%>%
  select(-NAME, -ends_with("M")) %>%
  rename(totalPop = B01003_001E,                           # Total Population
         medIncome = B19013_001E,                         # Median Household Income
         medRent = B25058_001E,                           # Median Rent
         ownerUnits = B25008_002E,                        # Owner-Occupied Units
         renterUnits = B25008_003E,                       # Renter-Occupied Units
         sameHouse75 = B07001_032E,                      # Same House 75 Years Ago
         sameHouse1 = B07001_017E,                       # Same House 1 Year Ago
         medianNoMortgage = B25088_003E,          # Median Selected Monthly Owner Costs (homes without a mortgage)
         medianMortgage = B25088_002E,        # Median Selected Monthly Owner Costs (homes with a mortgage)
         bachelors = B15003_022E,                      # Educational Attainment: Bachelor's Degree
         belowPoverty = B17001_002E,                  # Percentage of Population Below the Poverty Level
         highSpeedInternet = B28002_004E,  # Percentage of Housing Units with High-Speed Internet
         noVehicle = B25044_003E,           # Percentage of Housing Units with No Vehicle Available
         White = B02001_002E,                         # Race and Ethnicity: White Alone
         Black = B02001_003E,                         # Race and Ethnicity: Black or African American Alone
         hispanicLatino = B03001_003E                     # Race and Ethnicity: Hispanic or Latino
         )

# Transform the data to ESRI:102728 projection
kcTracts <- kcTracts %>% st_transform(st_crs(seattleData)) #need to add projection to kc stuff earlier

#also need to filter out the tracts for seattle 

```

```{r}
seattleSchools
```





