---
title: "Tidycensus Markdown"
author: "Revathi V. Machan"
date: "2024-01-26"
output: 
html_document:
  toc: yes
  toc_float: yes
  code_folding: hide
    
---

Week 2 PPA Assignment
```{r setup,include=FALSE, cache = TRUE}
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

if (!require(pacman)){install.packages("pacman"); library(pacman)}
install.packages("dplyr")
install.packages("tidycensus")
p_load (sf, tidyverse, knitr, kableExtra, rmarkdown, tidycensus, dplyr)
``` 

```{r insert census api key}
census_api_key("b3eda1fa84dde3c5ad443fd407d48f2584ab2726", overwrite = TRUE, install = TRUE)

```

```{r load census data dictionaries}
acsVariables2020 <- load_variables(2020, "acs5", cache = TRUE)
acsVariables2016 <- load_variables(2016, "acs5", cache = TRUE)
```

```{r downloading data from tidycensus}
acsVars <- c("B01001_001E", # ACS total Pop estimate
              "B25002_001E", # Estimate of total housing units
              "B25002_003E", # Number of vacant housing units
              "B19013_001E", # Median HH Income ($)
              "B02001_002E", # People describing themselves as "white alone"
              "B06009_006E") # Total graduate or professional degree)
```

```{r call api to get tract level data, cache = TRUE, message = FALSE, warning = FALSE}
acsTractsPhl2020 <- get_acs(geography = "tract",
                            year = 2020,
                            variables = acsVars,
                            geometry = FALSE,
                            state = "PA",
                            county = "Philadelphia",
                            output = "wide")
```

```{r wrangling data w dplyr, cache = TRUE}
acsTractsPhl2020 <- acsTractsPhl2020 %>%
  dplyr::select(GEOID, NAME,all_of(acsVars))

acsTractsPhl2020 <- acsTractsPhl2020 %>%
  rename(totalPop2020 = B01001_001E,
          totalHU2020 = B25002_001E,
          totalVacant2020 = B25002_003E,
          medHHIncome2020 = B19013_001E,
          totalWhite2020 = B02001_002E,
          totalGradDeg2020 = B06009_006E)

acsTractsPhl2020 <- acsTractsPhl2020 %>%
  mutate(pctVacancy2020 = totalVacant2020/totalHU2020,
         pctWhite = totalWhite2020/totalPop2020)
```

```{r 2015 acs, cache = TRUE, message = FALSE}
acsTractsPhl2016 <- get_acs(geography = "tract",
                            year= 2016,
                            variables = acsVars,
                            geometry = FALSE,
                            state = "PA",
                            county = "Philadelphia",
                            output = "wide") %>%
  dplyr::select(GEOID, NAME, all_of(acsVars)) %>%
  rename(totalPop2016 = B01001_001E,
          totalHU2016 = B25002_001E,
          totalVacant2016 = B25002_003E,
          medHHIncome2016 = B19013_001E,
          totalWhite2016 = B02001_002E,
          totalGradDeg2016 = B06009_006E) %>%
  mutate(pctVacancy2016 = totalVacant2016/totalHU2016,
         pctWhite2016 = totalWhite2016/totalPop2016)
```

```{r left join, cache = TRUE}
allACS <- left_join(acsTractsPhl2016, acsTractsPhl2020,
                    by= c("GEOID"))
```


```{r mutate data, cache = TRUE}
allACS <- allACS %>%
  mutate(changeMedHHIncome = medHHIncome2020 - (medHHIncome2016*1.08),
         changeGradDegreePct = (totalGradDeg2020/totalPop2020) - (totalGradDeg2016/totalPop2016)) 
```

```{r base summaries, cache = TRUE}
mean(allACS$changeMedHHIncome, na.rm = TRUE)
median(allACS$changeMedHHIncome, na.rm = TRUE)
```
```{r basic histogram, cache = TRUE}
hist(allACS$changeMedHHIncome)
```

```{r ggplot for histogram, warning = FALSE, cache = TRUE, message = FALSE}
ggplot(allACS) +
  geom_histogram(aes(changeMedHHIncome), binwidth = 5000)+
  labs(
    title = "Change in Philadelphia HH median income by tract, 2016-2020",
    caption = "Data: US Census Bureau, ACS 5-year estimates",
    x = "Change in Med HH Income (2020 dollars)",
    y="Number of tracts"  )
```
```{r my tracts, cache = TRUE}

myTracts <- c("42101023500", 
              "42101023600", 
              "42101023700", 
              "42101025300", 
              "42101025400",
              "42101025500", 
              "42101025600", 
              "42101038800")

allACS <- allACS %>%
  mutate(mtAiry = ifelse(GEOID %in% myTracts, "MT AIRY", "REST OF PHILADELPHIA"))

```


The following plot is a city-wide plot that shows the change in raw housing units from 2016 to 2020, with the Mount Airy vacant housing numbers depicted in a separate color
```{r ggplot city wide, cache = TRUE}
allACS <- allACS %>%
  mutate(changeMedVacantHousing = totalVacant2020 - (totalVacant2016))

ggplot(allACS)+
  geom_point(aes(x=totalVacant2016,
                 y=totalVacant2020,
                 color = mtAiry))+
  labs(
    title = "2020 Vacant Housing as a Function of 2016 Vacant Housing",
    subtitle = "Revathi V. Machan",
    caption = "Data: US Census Bureau, ACs 5-year estimates",
    x="Vacant Housing 2016",
    y="Vacant Housing 2020")

```

 
The following map shows the number of vacant housing units in 2016 as a percentage of total housing units on a per-tract basis, showing that there is a lower percentage of vacancy in Mount airy than in adjacent tracts.
```{r 2020tractsSf, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
acsTractsPHL2016Sf <- get_acs(geography = "tract",
                             year = 2016, 
                             variables = acsVars, 
                             geometry = TRUE, 
                             state = "PA", 
                             county = "Philadelphia", 
                             output = "wide") %>% 
  dplyr::select (GEOID, NAME, all_of(acsVars)) %>%
  rename (totalPop2016 = B01001_001E,
          totalHU2016 = B25002_001E,
          totalVacant2016 = B25002_003E,
          medHHIncome2016 = B19013_001E,
          totalWhite2016 = B02001_002E,
          totalGradDeg2016 = B06009_006E) %>%
  mutate(vacancyHU2016 = totalVacant2016/totalHU2016) %>%
  mutate(mtAiry = ifelse(GEOID %in% myTracts, "MT AIRY", "REST OF PHILADELPHIA"))

acsTractsPHL2020Sf <- get_acs(geography = "tract",
                             year = 2020, 
                             variables = acsVars, 
                             geometry = TRUE, 
                             state = "PA", 
                             county = "Philadelphia", 
                             output = "wide") %>% 
  dplyr::select (GEOID, NAME, all_of(acsVars)) %>%
  rename (totalPop2020 = B01001_001E,
          totalHU2020 = B25002_001E,
          totalVacant2020 = B25002_003E,
          medHHIncome2020 = B19013_001E,
          totalWhite2020 = B02001_002E,
          totalGradDeg2020 = B06009_006E) %>%
  mutate(mtAiry = ifelse(GEOID %in% myTracts, "MT AIRY", "REST OF PHILADELPHIA"))
```


```{r ggplot_geom_sf, warning = FALSE, cache = TRUE}
ggplot()+
  geom_sf(data = acsTractsPHL2016Sf, aes(fill = vacancyHU2016),
          color = "transparent")+
  geom_sf(data = acsTractsPHL2016Sf %>%
            filter(mtAiry == "MT AIRY") %>%
            st_union(),
          color = "white",
          fill = "transparent")+
  labs(
    title = "2016 Vacant Housing Units as a function of Total Housing Units",
    subtitle = "",
    caption = "Data: US Census Bureau, ACS 5-year estimates")+
  theme_minimal()

```

The following table shows a summary of mean vacant housing units per tract for Mount Airy versus the rest of Philadelphia as a whole in 2020.
```{r kable table}
vacancySumTable <- allACS %>%
  group_by(mtAiry) %>%
  summarize(meanVacancyHU = mean(totalVacant2020, na.rm = TRUE))

vacancySumTable %>%
  kbl%>%
  kable_styling(bootstrap_options = "striped", font_size = 9)
```
